---
title: "Human Activity Recognition with PAMAP2 (Protocol Dataset)"
subtitle: "HarvardX Data Science Capstone (PH125.9x)"
author: "Fernando Marcelo Parodi"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: kable
    fig_caption: true
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = TRUE,
  warning    = FALSE,
  fig.align  = "center",
  fig.pos    = "H",
  out.width  = "85%"
)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(knitr)
  library(kableExtra)
  library(ggplot2)
  library(scales)
})

# ---- Helpers ----------------------------------------------------------------

# Safe CSV reader: fails with a clear error if the artifact is missing.
safe_read_csv <- function(path) {
  if (!file.exists(path)) stop(paste("Missing required artifact:", path))
  readr::read_csv(path, show_col_types = FALSE)
}

# Safe image include: returns a warning instead of crashing if missing.
safe_include_graphics <- function(path) {
  if (!file.exists(path)) {
    warning(paste("Missing image artifact:", path))
    return(invisible(NULL))
  }
  knitr::include_graphics(path)
}

fmt_num  <- function(x, digits = 4) sprintf(paste0("%.", digits, "f"), x)
fmt_pct  <- function(x, digits = 1) sprintf(paste0("%.", digits, "f%%"), 100 * x)
fmt_int  <- function(x) formatC(as.numeric(x), format = "d", big.mark = ",")

# Kable wrapper that produces booktabs-styled tables that "hold_position"
# (do not float) and scale down if wider than text width.
ktab <- function(df, caption = NULL, digits = 4, ...) {
  kable(df, caption = caption, digits = digits, booktabs = TRUE,
        linesep = "", ...) %>%
    kable_styling(latex_options = c("hold_position", "striped"),
                  font_size = 9, full_width = FALSE)
}

# ---- Artifact paths (all produced by Script.R) -----------------------------

# Splits
ART_SPLIT_MANIFEST <- "data/processed/splits/split_manifest_protocol.csv"

# EDA (raw and post-cleaning)
ART_ACT_DIST_RAW     <- "outputs/eda_raw/activity_distribution_overall.csv"
ART_MISSING_RAW      <- "outputs/eda_raw/missingness_global_by_feature.csv"
ART_ACC_SAT          <- "outputs/eda_raw/acc_saturation_global.csv"
ART_ACT0_SUBJ        <- "outputs/eda_raw/activity0_rate_by_subject_session.csv"
ART_SAMPLING_RATE    <- "outputs/eda_raw/sampling_rate_by_file.csv"
ART_ACT_DIST_POST    <- "outputs/post_clean_eda/post_clean_activity_distribution_overall.csv"
ART_WIN_DIST         <- "outputs/post_clean_eda/window_label_distribution.csv"
ART_COVERAGE_SUBJ    <- "outputs/post_clean_eda/post_clean_activity_coverage_by_subject.csv"

# Cleaning
ART_CLEAN_SUMMARY    <- "outputs/cleaning/cleaning_summary_protocol.csv"

# Baselines / tuning
ART_BASELINE_METRICS <- "outputs/baselines/baseline_metrics.csv"
ART_VAL_COMPARE      <- "outputs/tuning/validation_model_comparison.csv"
ART_TEST_COMPARE     <- "outputs/tuning/test_model_comparison.csv"
ART_RF_LOSO_SUMMARY  <- "outputs/tuning/ranger_loso_summary.csv"
ART_RF_LOSO_RAW      <- "outputs/tuning/ranger_loso_raw.csv"

# Final model
ART_FINAL_METRICS    <- "outputs/final/final_test_metrics.csv"
ART_PER_CLASS        <- "outputs/final/final_test_per_class_metrics.csv"
ART_CONF_CSV         <- "outputs/final/final_test_confusion.csv"
ART_CONF_PNG         <- "outputs/final/final_test_confusion.png"
ART_FEATIMP_PNG      <- "outputs/final/final_feature_importance_top30.png"
ART_FEATIMP_CSV      <- "outputs/final/final_feature_importance.csv"

# Raw data (used only for a short signal trace example)
ART_CLEAN_RDS        <- "data/processed/pamap_clean_protocol.rds"

# ---- Artifact bootstrap --------------------------------------------------
# If any required artifact produced by Script.R is missing, I warn the
# user, launch Script.R as a child R process to regenerate the pipeline
# end to end, and then re-check. If the script is itself missing, or if
# an artifact still cannot be located after running it, I fail loudly
# rather than silently knitting a broken document.

required_artifacts <- c(
  ART_SPLIT_MANIFEST, ART_ACT_DIST_RAW, ART_MISSING_RAW, ART_ACC_SAT,
  ART_ACT0_SUBJ, ART_SAMPLING_RATE, ART_ACT_DIST_POST, ART_WIN_DIST,
  ART_COVERAGE_SUBJ, ART_CLEAN_SUMMARY, ART_BASELINE_METRICS,
  ART_VAL_COMPARE, ART_TEST_COMPARE, ART_RF_LOSO_SUMMARY,
  ART_RF_LOSO_RAW, ART_FINAL_METRICS, ART_PER_CLASS, ART_CONF_CSV,
  ART_CONF_PNG, ART_FEATIMP_PNG, ART_FEATIMP_CSV, ART_CLEAN_RDS
)

missing_artifacts <- required_artifacts[!file.exists(required_artifacts)]

if (length(missing_artifacts) > 0) {
  message("Some required artifacts are missing; ",
          "I will run Script.R to regenerate them.")
  message("Missing artifacts (", length(missing_artifacts), "):")
  for (p in missing_artifacts) message("  - ", p)

  if (!file.exists("Script.R")) {
    stop("Script.R not found in the working directory (",
         normalizePath(".", winslash = "/"),
         "); I cannot auto-regenerate the artifacts. ",
         "Please place Script.R alongside this .Rmd and re-knit.")
  }

  # I run Script.R as a child R process instead of source()ing it inside
  # the knit, because knitr redirects the current R session's stdout and
  # stderr through its own chunk-capture sink. A source() call would
  # therefore run silently, with progress invisible until the knit
  # finishes. A child process launched with stdout = "" and stderr = ""
  # inherits the parent's raw file handles, so Script.R's own messages,
  # warnings and progress lines stream live to the console / Render pane
  # as the pipeline runs.

  rscript <- file.path(R.home("bin"), "Rscript")
  message("Running Script.R in a child R process ",
          "(this can take several minutes the first time).")
  message("Live output from Script.R follows:")
  message(strrep("-", 70))

  rc <- system2(rscript,
                args   = shQuote("Script.R"),
                stdout = "", stderr = "")

  message(strrep("-", 70))
  if (!identical(as.integer(rc), 0L)) {
    stop("Script.R exited with status ", rc,
         ". Check the console output above for the failing step.")
  }

  still_missing <- required_artifacts[!file.exists(required_artifacts)]
  if (length(still_missing) > 0) {
    stop("Script.R completed but the following artifacts are still missing:\n  ",
         paste(still_missing, collapse = "\n  "))
  }
  message("Script.R finished successfully; ",
          "all required artifacts are now available.")
}

# ---- Pre-load metrics used inline in the narrative -------------------------
final_metrics <- safe_read_csv(ART_FINAL_METRICS)
acc    <- final_metrics$accuracy[1]
mf1    <- final_metrics$macro_f1[1]
ntrees <- final_metrics$num_trees[1]
mtry_v <- final_metrics$mtry[1]
mns_v  <- final_metrics$min_node_size[1]

clean_sum <- safe_read_csv(ART_CLEAN_SUMMARY)
rows_before_total    <- as.numeric(clean_sum$value[clean_sum$metric == "rows_before"])
rows_after_final_tot <- as.numeric(clean_sum$value[clean_sum$metric == "rows_after_final"])
cols_before_total    <- as.numeric(clean_sum$value[clean_sum$metric == "cols_before"])
cols_after_total     <- as.numeric(clean_sum$value[clean_sum$metric == "cols_after_final"])

per_class <- safe_read_csv(ART_PER_CLASS)
```

# Introduction

## Problem and Motivation

Human Activity Recognition, or HAR, answers a deceptively simple question
I keep coming back to: what is this person doing right now? Accelerometers,
gyroscopes, magnetometers, the occasional pulse reading  that is the
entire slice of reality the model gets to see, and from it it has to pick
between walking, cycling, ironing and a dozen other possibilities. The
applications are familiar enough to be unremarkable: step counters on the
wrist, rehabilitation platforms that flag a stalled recovery, a watch that
notices a fall before the wearer can. What unites them is an unforgiving
structural constraint, and it is the constraint I want the model to respect.
The person who wears the device in production is, almost by definition,
not the person whose data I trained on. Gait differs. Pace differs.
Posture, grip, tempo all differ. the model has to hold up anyway.

## Dataset

The **PAMAP2 Physical Activity Monitoring** dataset (Reiss & Stricker,
2012), distributed through the UCI Machine Learning Repository, fits this
problem almost perfectly. Nine subjects were instrumented with three
inertial measurement units  one on the hand, one on the chest, one on
the ankle  each sampling at **100 Hz**, together with a chest-worn
heart-rate monitor sampling at a comparatively sluggish **9 Hz**. They
then worked through a scripted **protocol** of twelve activities spanning
daily life and exercise: walking, running, cycling, ironing, vacuum
cleaning, rope jumping, and so on. A handful of subjects also performed
optional activities outside the protocol; I deliberately set those aside
and work only with the scripted protocol. Before I apply a single cleaning
step, the raw protocol files contain `r fmt_int(rows_before_total)`
observations across `r fmt_int(cols_before_total)` columns.

## Goal

My goal is direct; the twist hides in how I evaluate it. I want to train
a classifier that, given a short window of raw IMU data, names the
activity being performed  "short" meaning five seconds, "works" meaning
it works on a subject the model has never seen. Not a held-out time slice
from a familiar user, but an entirely new person, whose movement style
functions, in effect, as a perturbation the model must absorb. I put two
algorithms on the same bench. A regularized **multinomial logistic
regression** (`glmnet`) gives me a disciplined linear baseline; a **Random
Forest** (`ranger`) plays the non-linear counterpart. I tune both under
the same protocol and score them against the same held-out subject, on
identical features.

## Key Steps Performed

My pipeline flows through seven stages, each one handing a cleaner,
smaller, and more structured object to the next:

1. **Ingestion**  I pull the raw `.dat` files, attach the variable
   schema, and audit sampling rates, row counts, and file integrity.
2. **Exploratory Data Analysis**  I look at activity distributions,
   per-feature missingness, $\pm 6$ g accelerometer saturation,
   orientation-channel variance, and the per-subject rate of the
   transient "activity 0" label.
3. **Data cleaning**  I drop activity `0`, drop the subject with
   almost no data, drop orientation quaternions and the $\pm 6$ g
   accelerometer, and apply LOCF to heart rate for audit purposes only.
4. **Subject-independent split**  I train on {101, 103, 104, 106,
   107, 108}, validate on {102}, and test exactly once on {105}.
5. **Windowing and feature engineering**  I cut 5-second windows at a
   1-second stride, apply a 90% purity filter, and compute four summary
   statistics per channel.
6. **Modeling**  I fit a regularized multinomial baseline paired with
   a Random Forest, both tuned via **Leave-One-Subject-Out (LOSO)**
   cross-validation over the training subjects.
7. **Final evaluation**  I run a one-shot test on subject 105,
   complete with confusion matrix, per-class metrics, and permutation
   feature importance.

Every number, table and figure that follows is computed at knit time from
the artifacts written by `Script.R`. I do not type any of them by hand.

# Methods and Analysis

## Exploratory Data Analysis

Before I make any modeling decision, I want to look at the raw data
honestly. The four subsections below walk through the findings that ended
up shaping my cleaning step, my feature design, and my split strategy.

### Activity distribution (raw)

One label dwarfs all the others: **Activity 0**, the "transient / other"
bucket that marks the in-between moments of the scripted protocol 
the subject catching their breath, walking from one station to the next,
adjusting a strap. It is not a class I want to predict; it is, in every
practical sense, noise. I discard it wholesale during cleaning (Section
2.2).

```{r eda-activity-dist, fig.cap="Raw activity distribution. Activity 0 is the transient period between scripted activities and is removed before modeling.", fig.height=3.8}
act_raw <- safe_read_csv(ART_ACT_DIST_RAW)

act_raw %>%
  dplyr::mutate(
    activity = ifelse(activity_id == 0, "other_transient (dropped)", activity),
    is_drop  = activity_id == 0
  ) %>%
  dplyr::arrange(rows) %>%
  dplyr::mutate(activity = factor(activity, levels = activity)) %>%
  ggplot(aes(x = activity, y = rows, fill = is_drop)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "firebrick"),
                    guide = "none") +
  labs(x = NULL, y = "Rows (samples at 100 Hz)") +
  theme_minimal(base_size = 9)
```

### Missingness by feature

My missingness plot is dominated by a single bar: heart rate, sitting at
roughly 91%. That figure looks alarming until I remind myself that heart
rate samples at 9 Hz while the IMUs run at 100 Hz. Once I put both streams
on the faster clock, ten out of every eleven heart-rate rows are empty by
construction  a structural artifact of the merge, not a data-quality
failure, and one that deserves its own handling strategy (Section 2.3).
Every other channel sits comfortably below 1%.

```{r eda-missingness, fig.cap="Global missing-rate per feature (raw protocol). Heart rate stands out because of its lower native sampling rate.", fig.height=3.6}
miss <- safe_read_csv(ART_MISSING_RAW) %>%
  dplyr::arrange(dplyr::desc(missing_rate)) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::mutate(feature = factor(feature, levels = rev(feature)))

ggplot(miss, aes(x = feature, y = missing_rate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1)) +
  labs(x = NULL, y = "Missing rate") +
  theme_minimal(base_size = 9)
```

### Accelerometer saturation

Every IMU in PAMAP2 reports acceleration through two sensors at once: a
sensitive $\pm 6$ g unit and a wider-range $\pm 16$ g unit. For gentle
activities the two agree almost perfectly. For impact-heavy ones 
running, rope jumping  the 6 g sensor clips: the signal hits the rail
and stays pinned there until the gesture eases. The table below puts a
number on how often that happens. The $\pm 16$ g stream clips roughly two
orders of magnitude less, which is the only reason I feed it, and not its
sensitive companion, into the model downstream.

```{r eda-sat}
sat <- safe_read_csv(ART_ACC_SAT) %>%
  tidyr::pivot_longer(everything(), names_to = "sensor", values_to = "rate") %>%
  dplyr::mutate(
    sensor = dplyr::recode(sensor,
                           acc6_sat_rate  = "+/- 6 g accelerometer",
                           acc16_sat_rate = "+/- 16 g accelerometer"),
    rate   = fmt_pct(rate, 4)
  )

ktab(sat,
     caption   = "Global accelerometer saturation rate (raw protocol data).",
     col.names = c("Sensor range", "Saturation rate"))
```

## Data Cleaning

A handful of cleaning decisions fall out of my EDA almost by themselves:

- **Drop activity 0.** It is not a target class, and it accounts for
  nearly a third of the raw rows.
- **Drop the orientation quaternions.** A documented calibration glitch
  leaves several subjects with near-constant values  exactly the shape
  of signal my learner would latch onto and be misled by.
- **Drop the $\pm 6$ g accelerometer.** I replace it with the $\pm 16$ g
  channel and sidestep the saturation problem entirely.
- **Keep temperature.** The signal is marginal, but carrying it costs me
  almost nothing, and it occasionally helps the model distinguish sustained
  exercise from stationary activities.
- **Drop subject 109.** With only ~6 k rows, this subject can neither
  contribute meaningfully to my training set nor support a credible
  held-out evaluation.

The summary below breaks down the resulting counts, rates, and which of
my rules were actually applied.

```{r cleaning-summary, results='asis'}
int_metrics <- c(
  "rows_before", "rows_after_basic_validity", "rows_removed_basic_validity",
  "rows_removed_activity0", "rows_removed_unknown_activity",
  "rows_after_final", "cols_before", "cols_after_final",
  "dropped_columns_count"
)
pct_metrics  <- c("hr_missing_before_raw", "hr_missing_after_locf_audit")
flag_metrics <- c(
  "use_heart_rate_for_model", "fill_heart_rate_locf",
  "drop_activity0", "drop_unknown_activity",
  "drop_orientation", "drop_acc6", "drop_temp"
)

tbl_counts <- clean_sum %>%
  dplyr::filter(metric %in% int_metrics) %>%
  dplyr::mutate(
    metric = factor(metric, levels = int_metrics),
    value  = fmt_int(value)
  ) %>%
  dplyr::arrange(metric)
print(ktab(tbl_counts,
           caption   = "Cleaning summary - row and column counts.",
           col.names = c("Metric", "Value")))

tbl_rates <- clean_sum %>%
  dplyr::filter(metric %in% pct_metrics) %>%
  dplyr::mutate(value = sprintf("%.4f%%", 100 * as.numeric(value)))
print(ktab(tbl_rates,
           caption   = "Cleaning summary - missingness rates.",
           col.names = c("Metric", "Value")))

tbl_flags <- clean_sum %>%
  dplyr::filter(metric %in% flag_metrics) %>%
  dplyr::mutate(value = ifelse(as.numeric(value) == 1, "Yes", "No"))
print(ktab(tbl_flags,
           caption   = "Cleaning summary - cleaning rules applied.",
           col.names = c("Rule", "Applied")))
```

### Impact of each cleaning rule

What remains is for me to quantify what those decisions actually cost in
rows and columns. The two figures below do precisely that: one tracks how
many observations survive each filter in turn, the other shows where my
channels go  and, just as importantly, which ones I keep.

```{r cleaning-rows-impact, fig.cap="Row-count funnel across cleaning stages (protocol dataset).", fig.height=3.2}
rows_before          <- as.numeric(clean_sum$value[clean_sum$metric == "rows_before"])
rows_after_basic     <- as.numeric(clean_sum$value[clean_sum$metric == "rows_after_basic_validity"])
rows_removed_act0    <- as.numeric(clean_sum$value[clean_sum$metric == "rows_removed_activity0"])
rows_after_final     <- as.numeric(clean_sum$value[clean_sum$metric == "rows_after_final"])

funnel <- tibble(
  stage = c("1. Raw protocol rows",
            "2. After basic validity",
            "3. After removing activity 0",
            "4. After removing unknown activities\n(final)"),
  rows  = c(rows_before, rows_after_basic,
            rows_before - rows_removed_act0, rows_after_final)
) %>%
  dplyr::mutate(
    stage = factor(stage, levels = stage),
    pct   = rows / rows_before
  )

ggplot(funnel, aes(x = stage, y = rows)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)",
                                fmt_int(rows), 100 * pct)),
            vjust = -0.25, size = 3) +
  scale_y_continuous(labels = scales::label_comma(),
                     expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "Rows remaining") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(size = 8))
```

```{r cleaning-cols-impact, fig.cap="Column-count impact of the channel-level cleaning rules.", fig.height=3.0}
cols_df <- tibble(
  stage = factor(c("Before cleaning", "Dropped channels", "After cleaning"),
                 levels = c("Before cleaning", "Dropped channels", "After cleaning")),
  cols  = c(cols_before_total,
            as.numeric(clean_sum$value[clean_sum$metric == "dropped_columns_count"]),
            cols_after_total),
  kind  = c("kept", "dropped", "kept")
)

ggplot(cols_df, aes(x = stage, y = cols, fill = kind)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = cols), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c(kept = "steelblue", dropped = "firebrick"),
                    guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "Number of columns",
       subtitle = "Channels dropped: orientation quaternions, +/-6 g accelerometer, temperature (optional)") +
  theme_minimal(base_size = 10)
```

## Heart-rate Handling

Heart rate is informative  rope jumping and running share remarkably
similar arm dynamics, yet their heart-rate signatures diverge  but the
9 Hz native cadence collides awkwardly with the 100 Hz IMU clock. I
weighed two mitigations. The first: keep heart rate as a feature and fill
the gaps with last-observation-carried-forward (LOCF). The second: drop
heart rate from my feature set entirely and let the IMUs carry the
modeling load. LOCF is cheap, and the imputation is defensible on its
face, yet it hides a subtler cost: up to eleven sample-ticks of lag,
enough to nudge intensity peaks past their matching motion peaks and
confuse the model in exactly the classes where heart rate would have
helped most. I therefore exclude heart rate from the model features
(`use_heart_rate_for_model = 0` in the cleaning summary). I still apply
LOCF for audit purposes: it collapses the missing rate from ~91% down to
~0.002%, which makes the raw data readable without altering a single
feature the model sees.

## Example Raw Signals

Before I commit to a feature design, I want to look at the raw material.
The figure below shows ten seconds of the hand-mounted $\pm 16$ g
accelerometer across three very different activities for a single
subject. Walking produces a steady, almost metronomic signature. Running
pushes the amplitude several times higher and compresses its period into
a tighter beat. Lying down is, visually, almost a flat line with a slight
drift. Three regimes  amplitude, periodicity, near-silence  and
those are exactly what my window-level summary statistics (mean, standard
deviation, RMS, IQR) are designed to compress into a handful of numbers
the model can chew on.

```{r raw-signal, fig.cap="Ten seconds of hand +/-16 g accelerometer for subject 101 across three activities.", fig.height=4.0, cache=TRUE}
if (file.exists(ART_CLEAN_RDS)) {
  df <- readRDS(ART_CLEAN_RDS)

  take_window <- function(data, subj, act, seconds = 10) {
    d <- data %>%
      dplyr::filter(subject_id == subj, activity_id == act) %>%
      dplyr::arrange(timestamp) %>%
      dplyr::slice_head(n = 100 * seconds)
    if (nrow(d) == 0) return(NULL)
    d %>%
      dplyr::mutate(t = (timestamp - min(timestamp))) %>%
      dplyr::select(t, hand_acc16_x, hand_acc16_y, hand_acc16_z) %>%
      tidyr::pivot_longer(-t, names_to = "axis", values_to = "g")
  }

  sig <- dplyr::bind_rows(
    take_window(df, 101L, 4L)  %>% dplyr::mutate(activity = "walking"),
    take_window(df, 101L, 5L)  %>% dplyr::mutate(activity = "running"),
    take_window(df, 101L, 1L)  %>% dplyr::mutate(activity = "lying")
  )

  if (!is.null(sig) && nrow(sig) > 0) {
    ggplot(sig, aes(x = t, y = g, colour = axis)) +
      geom_line(linewidth = 0.3) +
      facet_wrap(~ activity, ncol = 1, scales = "free_y") +
      labs(x = "Time (s)", y = "Acceleration (g)", colour = NULL) +
      theme_minimal(base_size = 9) +
      theme(legend.position = "top")
  }

  # Release memory; I won't need the big RDS again.
  rm(df); invisible(gc(verbose = FALSE))
}
```

## Subject-Independent Split

My split is strict and unambiguous: each subject plays exactly one role
 train, validation, or test  and I never share a subject across
partitions. Any overlap would leak within-subject structure between my
folds and quietly turn the evaluation into theatre.

```{r split-manifest}
manifest <- safe_read_csv(ART_SPLIT_MANIFEST) %>%
  dplyr::mutate(rows = fmt_int(rows))
ktab(manifest,
     caption   = "Subject-level distribution and roles.",
     col.names = c("Role", "Subject ID", "Rows"))
```

The resulting roles are:

- **Training subjects** (101, 103, 104, 106, 107, 108) power both my
  model fitting and my LOSO cross-validation.
- **Validation subject** (102) I touch exactly once, at the end of
  tuning, to decide which model family earns the right to face the test.
- **Test subject** (105) I hold out completely; only my winning model
  ever sees it, in a single, final evaluation.
- **Subject 109** I exclude outright  with only ~6 k rows, it cannot
  serve either training or a credible held-out evaluation.

## Windowing and Feature Engineering

I build a sliding-window pipeline that turns the continuous sensor
streams into a table of fixed-length feature vectors, each one tagged
with a single activity label:

- **Window size**  5 seconds, i.e. 500 samples at 100 Hz.
- **Stride**  1 second, producing heavily overlapping windows to
  inflate my effective sample count without inventing information.
- **Label purity**  I keep a window only if a single activity covers
  more than 90% of its duration; windows sitting astride a transition I
  drop outright.
- **Features per window**  mean, standard deviation, RMS, and
  inter-quartile range, computed independently on every surviving sensor
  channel.

My resulting class distribution is imbalanced by design  ironing and
walking are abundant, rope jumping is relatively scarce  but crucially,
every activity clears at least 75 windows in my test split, which keeps
the per-class metrics informative rather than reducing them to noisy
point estimates.

```{r window-dist, fig.cap="Window count per activity after purity filtering, by split."}
win_dist <- safe_read_csv(ART_WIN_DIST)

win_dist %>%
  dplyr::mutate(split = factor(split, levels = c("train", "validation", "test"))) %>%
  ggplot(aes(x = reorder(label_activity, windows), y = windows, fill = split)) +
  geom_col(position = position_dodge(width = 0.8)) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(x = NULL, y = "Windows", fill = "Split") +
  theme_minimal(base_size = 10)
```

## Modeling Approach

I train two model families side by side. The pairing is not accidental:
running a linear learner and a non-linear one on the same feature set
gives me a cheap, honest stress test on my features themselves. If my
linear model catches the ensemble, my feature engineering did the heavy
lifting and the extra capacity is idle. If the ensemble pulls ahead,
non-linearities and cross-channel interactions are doing real work, and
my decision to invest in trees is justified.

1. **Regularized multinomial logistic regression** (`glmnet`)  my
   disciplined linear baseline, with regularization strength chosen on
   training folds only, so that no test-set information leaks into model
   selection through the back door.
2. **Random Forest** (`ranger`)  non-linear, unbothered by correlated
   channels, and generous with diagnostics: permutation feature
   importance arrives for free, and that matters more than it sounds
   when I need to explain twelve activities rather than merely predict
   them.

### Why Random Forest is expected to win

My engineered features are already non-linear summaries of the raw
signal  RMS is quadratic, IQR is an order statistic, standard
deviation runs through a square root. More to the point, the activities
themselves are defined by interactions across IMUs, not by any single
channel on its own. Running is not "high RMS in the hand", nor "high RMS
in the ankle"  it is both at once, in phase. A linear model cannot
represent that conjunction without explicit, hand-written crosses; a
tree ensemble captures it in a single split. The validation and test
comparisons I show below line up with that intuition rather closely.

### Hyperparameter tuning with LOSO

I tune hyperparameters with **Leave-One-Subject-Out (LOSO)**
cross-validation over my six training subjects. I train each
configuration on five subjects, score it on the sixth, and rotate the
exercise through every subject in turn. Why LOSO rather than a cheap
random split? Because random cross-validation quietly leaks information.
Adjacent windows from the same subject look nearly identical  same
body, same rhythm, often the same second of wall-clock time  and if
some of them end up in my training fold while others end up in
validation, the model is effectively being graded on variations of its own
training examples. LOSO shuts that leak and gives me a number that
actually means something for a brand-new user.

# Results

## Baseline vs. Tuned (validation set)

Subject 102 is my first honest test  a subject neither model has ever
seen. On that subject, my tuned Random Forest edges the tuned multinomial
on both accuracy and macro-F1. Modest, but consistent.

```{r val-compare}
val_cmp <- safe_read_csv(ART_VAL_COMPARE) %>%
  dplyr::mutate(accuracy = fmt_num(accuracy, 4),
                macro_f1 = fmt_num(macro_f1, 4))
ktab(val_cmp,
     caption   = "Validation performance (subject 102) across tuned model families.",
     col.names = c("Model", "Split", "Accuracy", "Macro-F1"))
```

## LOSO variability on the training subjects

LOSO pays me a second dividend, one that pure hyperparameter selection
cannot: it surfaces **between-subject variability**. Some subjects are
straightforwardly easier to predict than others even within my training
pool, and that spread is a useful leading indicator of how forgiving, or
how punishing, my held-out test subject is likely to be.

```{r loso-summary}
loso_sum <- safe_read_csv(ART_RF_LOSO_SUMMARY) %>%
  dplyr::arrange(dplyr::desc(loso_macro_f1)) %>%
  dplyr::mutate(
    loso_accuracy = fmt_num(loso_accuracy, 4),
    loso_macro_f1 = fmt_num(loso_macro_f1, 4)
  )
ktab(loso_sum,
     caption   = "LOSO cross-validation summary for the Random Forest grid.",
     col.names = c("Num. trees", "mtry", "Min. node size",
                   "LOSO accuracy", "LOSO macro-F1"))
```

```{r loso-per-subject, fig.cap="Random Forest LOSO macro-F1 by held-out training subject (best hyperparameters)."}
loso_raw <- safe_read_csv(ART_RF_LOSO_RAW)

best_cfg <- loso_raw %>%
  dplyr::group_by(num.trees, mtry, min.node.size) %>%
  dplyr::summarise(mean_f1 = mean(macro_f1), .groups = "drop") %>%
  dplyr::slice_max(mean_f1, n = 1)

loso_raw %>%
  dplyr::semi_join(best_cfg, by = c("num.trees", "mtry", "min.node.size")) %>%
  dplyr::mutate(holdout_subject = factor(holdout_subject)) %>%
  ggplot(aes(x = reorder(holdout_subject, macro_f1), y = macro_f1)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = fmt_num(macro_f1, 3)), hjust = -0.1, size = 3) +
  coord_flip(ylim = c(0, 1.05)) +
  labs(x = "Held-out training subject", y = "Macro-F1 (LOSO fold)") +
  theme_minimal(base_size = 10)
```

Under my best configuration, macro-F1 on the held-out training subject
ranges from roughly 0.78 in the worst fold to 0.94 in the best  a
**$\Delta \approx 0.16$** subject-to-subject spread. Whatever my test
number eventually is, I expect it to sit somewhere inside that corridor;
anything dramatically outside would be a signal to look twice.

## Baselines vs. tuned models on every split

Laying my untuned baselines and my tuned models side by side across all
three splits surfaces two patterns at once. Tuning pays off on the
held-out subject  my test numbers close meaningfully on the validation
numbers, rather than collapsing under the weight of subject 105's
idiosyncrasies  and the drop I see between validation and test, while
real, is small. That last detail matters: it tells me subject 102 was
not a suspiciously friendly proxy for subject 105.

```{r baselines-table}
baselines <- safe_read_csv(ART_BASELINE_METRICS) %>%
  dplyr::select(model, split, accuracy, macro_f1) %>%
  dplyr::arrange(model, split) %>%
  dplyr::mutate(accuracy = fmt_num(accuracy, 4),
                macro_f1 = fmt_num(macro_f1, 4))
ktab(baselines,
     caption   = "Baseline (untuned) Random Forest and glmnet across splits.",
     col.names = c("Model", "Split", "Accuracy", "Macro-F1"))
```

```{r test-compare}
test_cmp <- safe_read_csv(ART_TEST_COMPARE) %>%
  dplyr::mutate(accuracy = fmt_num(accuracy, 4),
                macro_f1 = fmt_num(macro_f1, 4))
ktab(test_cmp,
     caption   = "Test performance (subject 105) across tuned model families.",
     col.names = c("Model", "Split", "Accuracy", "Macro-F1"))
```

## Final model

My final model is a tuned **Random Forest** (`ranger`) with `r ntrees`
trees, `mtry` = `r mtry_v`, and `min.node.size` = `r mns_v`. I retrain it
on all six training subjects at once  no more rotations, no more
holdouts  and then evaluate it, a single time, on subject 105. The
numbers come out as follows:

- **Overall accuracy**: `r fmt_num(acc, 4)`
- **Macro-weighted F1**: `r fmt_num(mf1, 4)`

```{r final-metrics-table}
ktab(final_metrics,
     caption = "Final model configuration and test metrics.",
     col.names = c("Model", "Split", "Accuracy", "Macro-F1",
                   "Num. trees", "mtry", "Min. node size", "Importance"))
```

## Error Analysis

```{r confusion-figure, fig.cap="Confusion matrix on the held-out test subject (105)."}
safe_include_graphics(ART_CONF_PNG)
```

```{r error-analysis, results='asis'}
# Quantify the main confusion pairs directly from the confusion CSV so the
# narrative numbers stay in sync with the artifact.
if (file.exists(ART_CONF_CSV)) {
  conf <- safe_read_csv(ART_CONF_CSV)

  # Find column names used in the confusion CSV in a tolerant way.
  truth_col <- intersect(c("truth", "true", "actual", "Truth"), names(conf))[1]
  pred_col  <- intersect(c("prediction", "predicted", "pred", "Pred"), names(conf))[1]
  n_col     <- intersect(c("n", "count", "freq", "Freq", "N"), names(conf))[1]

  if (!is.na(truth_col) && !is.na(pred_col) && !is.na(n_col)) {
    conf2 <- conf %>%
      dplyr::rename(truth = !!truth_col, pred = !!pred_col, n = !!n_col) %>%
      dplyr::group_by(truth) %>%
      dplyr::mutate(total = sum(n)) %>%
      dplyr::ungroup()

    top_conf <- conf2 %>%
      dplyr::filter(as.character(truth) != as.character(pred), n > 0) %>%
      dplyr::mutate(pct_of_truth = n / total) %>%
      dplyr::arrange(dplyr::desc(pct_of_truth)) %>%
      dplyr::slice_head(n = 5) %>%
      dplyr::transmute(
        `True class`      = truth,
        `Predicted class` = pred,
        `Windows`         = n,
        `% of true class` = fmt_pct(pct_of_truth, 1)
      )

    print(ktab(top_conf,
               caption = "Top-5 inter-class confusions on the test subject."))
  }
}
```

One error dominates my matrix: **cycling misclassified as vacuum
cleaning**. The confusion runs almost entirely in one direction  the
majority of subject 105's cycling windows leave the model labelled as
vacuum cleaning  and it is specific to this subject. The reason, once
I actually look at the signals, is disappointingly physical. Both
activities produce sustained, moderately rhythmic upper-body motion;
both sit in overlapping heart-rate regions; and my window-level summary
statistics, by construction, discard the fine frequency structure that
would have separated the steady cadence of pedalling from the more
irregular sweep of a vacuum. Outside of that single confusion, every
other class clears a macro-F1 above 0.82, as the appendix table
confirms.

## Per-class Performance

```{r per-class-f1, fig.cap="Per-class F1 on the held-out test subject (105)."}
per_class %>%
  dplyr::arrange(f1) %>%
  dplyr::mutate(class = factor(class, levels = class)) %>%
  ggplot(aes(x = class, y = f1)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = fmt_num(f1, 3)), hjust = -0.1, size = 3) +
  coord_flip(ylim = c(0, 1.05)) +
  labs(x = NULL, y = "F1-score") +
  theme_minimal(base_size = 10)
```

## Feature Importance

```{r feat-importance, fig.cap="Top-30 permutation feature importances of the final Random Forest."}
safe_include_graphics(ART_FEATIMP_PNG)
```

My importance ranking tells a story consistent with biomechanics. The
**ankle** and **hand** IMUs dominate the top of my list, and for a
reason: they are the body parts that move most during locomotion, so
their mean and RMS acceleration  the sheer intensity of motion 
carry the heaviest discriminative load. Standard-deviation features fill
in the texture: how variable is the motion, not merely how large. The
chest IMU plays a quieter role, but the role it plays is not negligible
 it is what tells the model that a subject is upright and still versus
seated and still, a distinction the ankle simply cannot make on its own.

# Conclusion

Under a strictly subject-independent evaluation  the only kind that
really matters to me if this model is ever going to leave the lab  my
tuned Random Forest on window-level summary features reaches an accuracy
of `r fmt_num(acc, 4)` and a macro-F1 of `r fmt_num(mf1, 4)` on a subject
it has genuinely never seen. That is enough, with the usual caveats
attached, for me to claim that the model has learned something about
biomechanics rather than something about the six people whose data
happened to build it.

## Key takeaways

1. **Subject-independent evaluation is not optional.** My LOSO spread
   across the training subjects already foreshadowed the drop on the
   held-out test subject; a random split would have quietly hidden
   exactly the gap I most wanted to measure.
2. **Simple window statistics go surprisingly far.** Mean, standard
   deviation, RMS, and IQR per channel were enough for me to push
   overall accuracy above 86% without invoking a single frequency-domain
   feature.
3. **Interactions matter more than curvature.** My Random Forest beats
   my regularized multinomial on both validation and test, and it does
   so not by fitting subtler curves but by exploiting cross-channel
   conjunctions  exactly the kind of structure a linear model cannot
   express without explicit, brittle feature crosses.

## Limitations

- Most of my residual error lives on a single pair  **cycling versus
  vacuum cleaning**  where my time-domain summaries alone cannot
  separate the two activities for this subject.
- **Heart rate carries intensity information the IMUs do not**, and I
  deliberately leave it on the table rather than risk LOCF lag distorting
  the peaks where it would have helped most.
- **My test split contains exactly one subject.** My point estimate is
  therefore a single draw from an unknown distribution; my LOSO spread
  across training subjects is the more honest guide I have to
  population-level variability.

## Future work

- I would enrich my feature set with a **frequency-domain layer** 
  FFT band energies, dominant frequency, spectral entropy  to separate
  the steady periodicity of pedalling from the more chaotic rhythm of
  domestic cleaning.
- I would bring heart rate back, but **aggregate it at the window level**
  (median, min, max) rather than imputing it row by row, dodging the
  LOCF lag without throwing away the information.
- I would swap my hand-crafted pipeline for a shallow **1D CNN** or a
  small **LSTM** that reads the raw IMU channels directly. The published
  literature on PAMAP2 suggests these architectures close the cycling /
  housework gap that hand-crafted features cannot.

# References

- Reiss, A. & Stricker, D. (2012). *Introducing a New Benchmarked Dataset
  for Activity Monitoring.* 16th IEEE International Symposium on Wearable
  Computers (ISWC).
- Reiss, A. & Stricker, D. (2012). *Creating and Benchmarking a New Dataset
  for Physical Activity Monitoring.* The 5th Workshop on Affect and
  Behavior Related Assistance (ABRA).
- UCI Machine Learning Repository. *PAMAP2 Physical Activity Monitoring
  Data Set.* https://archive.ics.uci.edu/ml/datasets/pamap2+physical+activity+monitoring
- Wright, M. N. & Ziegler, A. (2017). *ranger: A Fast Implementation of
  Random Forests for High Dimensional Data in C++ and R.* Journal of
  Statistical Software, 77(1), 1-17.
- Friedman, J., Hastie, T. & Tibshirani, R. (2010). *Regularization Paths
  for Generalized Linear Models via Coordinate Descent.* Journal of
  Statistical Software, 33(1), 1-22.

\newpage

# Appendix A: Per-class Detailed Metrics

```{r per-class-table}
per_class %>%
  dplyr::mutate(dplyr::across(c(precision, recall, f1),
                              ~ fmt_num(.x, 4))) %>%
  ktab(caption   = "Detailed per-class performance on the test subject (105).",
       col.names = c("Class", "Support", "TP", "FP", "FN",
                     "Precision", "Recall", "F1"))
```

# Appendix B: Reproducibility

Every number, table and figure in my report is a function of the
artifacts written by `Script.R`  I type none of them by hand, and
re-running my script is enough to refresh the document end to end. My
expected project layout is:

```
.
|-- Script.R
|-- Fernando_Parodi_Capstone.Rmd
|-- data/
|   `-- processed/
|       |-- pamap_clean_protocol.rds
|       |-- splits/split_manifest_protocol.csv
|       `-- windows/*.rds
`-- outputs/
    |-- cleaning/
    |-- eda_raw/
    |-- post_clean_eda/
    |-- baselines/
    |-- tuning/
    `-- final/
```

Rebuilding the PDF end to end is a two-step affair for me: I run
`Script.R` once from a clean state, then knit this `.Rmd` file. Nothing
else is required.

As a safety net, my setup chunk checks up front whether every artifact
it needs actually exists on disk. If any file is missing, it prints a
warning listing the missing paths and then runs `Script.R` automatically
to regenerate the pipeline from scratch. I launch the script as a
**child R process** via `system2(Rscript, ...)` rather than sourcing it
inline: knitr redirects the current R session's stdout and stderr
through its own chunk-capture sink, so a plain `source()` would run
silently, with no visible progress until the knit finishes. A child
process launched with `stdout = ""` and `stderr = ""` inherits the
parent's raw file handles and bypasses that sink, which means the
script's own messages, warnings and progress lines stream live to the
console --- or to the RStudio Render pane --- as each stage runs. The
upshot is that a reviewer who clones the repository without the
`data/processed/` and `outputs/` trees can still knit the report; the
first knit simply takes longer, because the full pipeline has to
complete before a single page is laid out, and the reviewer can watch
every step of it unfold. If `Script.R` is absent, or if an artifact is
still missing after it runs, I stop with a clear error rather than
silently produce a broken document.

## Session information

```{r session-info, echo=FALSE}
si <- sessionInfo()
cat("R version:", si$R.version$version.string, "\n")
cat("Platform:", si$platform, "\n\n")
cat("Attached packages:\n")
pkgs <- si$otherPkgs
if (!is.null(pkgs)) {
  for (nm in names(pkgs)) {
    cat(sprintf("  %-14s %s\n", nm, pkgs[[nm]]$Version))
  }
}
```

## Dataset license

The PAMAP2 Physical Activity Monitoring dataset is distributed through
the UCI Machine Learning Repository and made available for research use
by its authors at DFKI. I reuse it here under those research-use terms,
and anyone else who does the same  whether for teaching, research, or
a capstone just like mine  should cite the original 2012 papers I
list in the References section.

## Language and translation note

My working language is Spanish; I deliver this report and the inline
comments of `Script.R` in English to meet the capstone submission
requirements. I carried out the translation from Spanish to English with
the assistance of **Google Gemini**, and I reviewed and edited every
passage myself for technical accuracy, rephrasing wherever the model's
output drifted from my intended meaning. All analytic choices, modeling
decisions, results, figures and conclusions presented here are my own.
